A Tidy Framework and Infrastructure to Systematically Assemble Spatio-temporal Indexes from Multivariate Data

### 2023-08-17 tidyindex.qmd
use j for variable in the subscript instead in notation

```{=tex}
\begin{equation}
\mathbf{x}(s;t) = \{x^1(s;t), x^2(s;t), \cdots, x^p(s;t)\} \qquad s \in D_s \subseteq \mathbb{R}^m, t \in D_t \subseteq \mathbb{R}^n 
\end{equation}
```
where:

-   $x^i(s, t)$ represents a variable of interest for example precipitation, $i = 1, \cdots, p$ and

@@

In what follows when geographic or temporal components of the $x(s,t)$ process are fixed we will be using suffix notation. For example, $x_{s}(t)$ represents the data for a fixed location $s$ as a function of time $t$. While $x_{t}(s)$ denotes the spatial varying proccess for a fixed $t$. An overview of the notation for pipeline input, operation, and output is present in @tbl-notation.

| No. | Module                  | Input                                                                  | Operation                                                                         | Output                                                                 |
|-|-------------------|-----------------|-------------------|----------------------|
| 1   | Temporal processing     | $x^i_{s}(t)$                                                           | $f[x^i_s(t)]$                                                                         | $\tilde{x}^i_s(t^\prime)$, where $t^\prime \in D_{t^\prime}$            |
| 2   | Spatial processing      | $x^i_t(s)$                                                             | $g[x^i_t(s)]$                                                                         | $x^i_t(s^\prime)$, where $s^\prime \in D_{s^\prime}$ ????                   |
| 3   | Variable transformation | $x^i(s; t)$                                                            | $T[x^i(s;t)]$                                                                       | $Tx^i(s;t)$ ????                                                       |
| 4   | Scaling                 | $x^i(s; t)$                                                            | $[x^i(s;t) - \alpha]/\gamma$                                                        | $Sx^i(s;t)$ ????                                                       |
| 5   | Dimension reduction     | $\mathbf{x}(s;t)$, where $\mathbf{x} \subseteq \mathbb{R}^q,\; q\leq p$ | $h[\mathbf{x}(s;t)]$                                                              | $\mathbf{y}(s;t)$, where $\mathbf{y} \subseteq \mathbb{R}^d$ and $d < q$ |
| 6   | Distribution fit        | $x^i(s; t)$                                                            | $F[x^i(s;t)]$, where $F$ is the CDF | $p^i(s;t)$, where $p(.) \in [0, 1]$ is a probability value                |
| 7   | Normalising             | $x^i(s; t)$                                                            | $\Phi^{-1}[x^i(s; t)]$                                                              | $z^i(s; t)$, where $z$ has a normal distribution                          |
| 8   | Benchmarking            | $x^i(s; t)$                                                            | $u[x^i(s;t)]$                                                                       | $b^i(s;t)$                                                               |
| 9   | Simplification          | $x^i(s; t)$                                                            | $v[x^i(s;t)]$                                                                            | $A^i(s;t) \in \{a_1, a_2, \cdots, a_j\}$                                 |

@@

The temporal processing module takes as input argument a single variable $x_s^i(t)$ at location $s$ as a function of time. In this step the original time series can be transformed or summarized into a new one via time aggregation, the transformation is represented by the function $f$, $\tilde{x}_s^i(t^\prime) = f[x^i_s(t)]$ where $t^\prime$ refers to the new temporal resolution after aggregation. An example of a temporal processing done in the computation of the Standardized Precipitation Index (SPI), consists of summing the monthly precipitation series over a rolling time window of size $k$. That is also known as the time scale. For SPI, the choice of the time scale $k$ is used to control the accumulation period for the water deficit, enabling the assessment of drought severity across various types (meteorological, agricultural, and hydrological).

<!-- Assuming an original time series indexed by $t= 1, 2, \cdots, T$, the aggregated precipitation series, indexed by $t^\prime = 1, 2, \cdots, T-k+1$, can be expressed as $x(t^\prime) = \sum_{t = t^\prime}^{t^\prime + k -1}x(t)$.  -->

@@

The spatial processing module takes a single variable with a fixed temporal dimension, $x_t(s)$, as input. This step transforms the variable from the original spatial dimension $s$ into the new dimension $s^\prime \in D_{s^\prime}$ through $x^\#(s;t) = g[x(s;t)]$. The change of spatial dimension allows for the alignment of variables collected from different measurements, such as in-situ stations and satellite imagery, or originating from different resolutions. This also includes the aggregation of variables into different levels, such as city, state, and countries scales.

@@

Variable transformation takes input of a single variable $x^i(s;t)$ and reshapes its distribution using the function $T[.]$ to produce $x^i(s;t)$. When a variable has a skewed distribution, transformations such as log, square root, or cubic root can adjust the distribution towards normality. For example, in Human Development Index (HDI), a logarithmic transformation is applied to the variable Gross National Income per capita (GNI), to reduce its impact on HDI, particularly for countries with high GNI values.

@@

Unlike variable transformation, scaling maintains the distributional shape of the variable. It includes techniques such as centering, z-score standardisation, and min-max standardisation and can be expressed as $[x^i(s;t) - \alpha]/\gamma$. In Human Development Index (HDI), the three dimensions (health, education, and economy) are converted into the same scale (0-1) using min-max standardisation.

@@

Distribution fit applies the cumulative distribution function $F$ of a distribution on the variable $x^i(s; t)$ to obtain the probability values $p^i(s;t) \in [0, 1]$. In SPEI, many distributions, including log-logistic, pearson III, lognormal, and general extreme distribution, are candidates for the aggregated series. Different fitting methods, and different goodness of fit tests may be used to compare the distribution choice on the index value.

@@

<!-- This step maps the univariate series into a different scale, typically for ease of comparison across regions. For example, a normal scale, \[0, 1\], or \[0, 100\] may be favored for reporting certain indexes. In drought indexes, i.e. SPI or SPEI, the quantiles from the fitted distribution are converted into the normal scale via the normal reverse CDF function: $\Phi^{-1}(.)$. Normalising is usually used at the end of the pipeline and its main difference from the scaling step is that here the change of scale also changes the distribution of the variable. While being commonly used, this step can get criticism from analysts for forcing the data into the decided scale, which can be either unnecessary or inaccurately exaggerate or downplay the outliers. Also, the use of a normal scale needs to be interpreted with caution. @fig-normalising illustrates the normal density not being directly proportional to its probability of occurrence. This is concerning, especially at the extreme values, since a small difference in the tail density can have magnitudes of difference in its probability of occurrence. -->

@@

Benchmark sets a value $b^i(s,t)$ for comparing against the original variable $x^i(s;t)$. This benchmark can be a fixed value consistently across space and time or determined by the data through the function $u[x^i(s;t)]$. Once a benchmark being set, observations can be highlighted for adjustments in other modules or can serve targets for monitoring and planning.

@@

Simplification takes a continuous variable $x^i(s;t)$ and categorises it into a discrete set $A^i(s;t) \in \{a_1, a_2, \cdots, a_j\}$ through a piecewise function,

```{=tex}
\begin{equation}
v[x_i(s;t)] = 
\begin{cases}
a_0 & C_1 \leq x^i(s; t) < C_0 \\
a_1 & C_2 \leq x^i(s; t) < C_1 \\
a_2 & C_3 \leq x^i(s; t) < C_2 \\
\cdots \\
a_z & C_z \leq x^i(s; t)
\end{cases}
\end{equation}
```

This is typically used at the end of the index pipeline to simplify the index to communicate to the public the severity of the concept of interest measured by the index. An example of simplification is to map the calculated SPI to four categories: mild, moderate, severe, and extreme drought.

@@

  1) Top-ranked countries with GGGI > 0.85,
  2) Countries ranked between 57 and 62, with GGGI values from 0.72 to 0.73, and
  3) Low-ranked countries with GGGI < 0.6.

>>>>

```{=tex}
\begin{equation}
\mathbf{x}(s;t) = \{x_1(s;t), x_2(s;t), \cdots, x_p(s;t)\} \qquad s \in D_s \subseteq \mathbb{R}^m, t \in D_t \subseteq \mathbb{R}^n 
\end{equation}
```
where:

-   $x_j(s, t)$ represents a variable of interest for example precipitation, $j = 1, \cdots, p$ and

@@

In what follows when geographic or temporal components of the $x_j(s,t)$ process are fixed we will be using suffix notation. For example, $x_{sj}(t)$ represents the data for a fixed location $s$ as a function of time $t$. While $x_{tj}(s)$ denotes the spatial varying process for a fixed $t$. An overview of the notation for pipeline input, operation, and output is present in @tbl-notation.

| No. | Module                  | Input                                                                   | Operation                           | Output                                                                        |
|--|---------------|----------|---------------|--------------------------|
| 1   | Temporal processing     | $x_{sj}(t)$                                                             | $f[x_{sj}(t)]$                      | $x^{\text{Temp}}_{sj}(t^\prime) \quad t^\prime \in D_{t^\prime}$ |
| 2   | Spatial processing      | $x_{tj}(s)$                                                             | $g[x_{tj}(s)]$                      | $x^{\text{Spat}}_{tj}(s^\prime) \quad s^\prime \in D_{s^\prime}$                 |
| 3   | Variable transformation | $x_{j}(s; t)$                                                           | $T[x_j(s;t)]$                       | $x^{\text{Trans}}_j(s;t)$                                                     |
| 4   | Scaling                 | $x_j(s; t)$                                                             | $[x_j(s;t) - \alpha]/\gamma$        | $x^{\text{Scale}}_j(s;t)$                                                     |
| 5   | Dimension reduction     | $\mathbf{x}(s;t) \quad \mathbf{x} \subseteq \mathbb{R}^q,\; q\leq p$ | $h[\mathbf{x}(s;t)]$                | $\mathbf{y}(s;t) \quad \mathbf{y} \subseteq \mathbb{R}^d, d < q$      |
| 6   | Distribution fit        | $x_j(s; t)$                                                             | $F[x_j(s;t)]$ | $p_j(s;t) \quad p(.) \in [0, 1]$                   |
| 7   | Normalising             | $x_j(s; t)$                                                             | $\Phi^{-1}[x_j(s; t)]$              | $z_j(s; t)$                              |
| 8   | Benchmarking            | $x_j(s; t)$                                                             | $u[x_j(s;t)]$                       | $b_j(s;t)$                                                                    |
| 9   | Simplification          | $x_j(s; t)$                                                             | $v[x_j(s;t)]$                       | $A_j(s;t) \in \{a_1, a_2, \cdots, a_j\}$                                      |

@@

The temporal processing module takes as input argument a single variable $x_{sj}(t)$ at location $s$ as a function of time. In this step the original time series can be transformed or summarized into a new one via time aggregation, the transformation is represented by the function $f$, $x^{\text{Temp}}_{sj}(t^\prime) = f[x_{sj}(t)]$ where $t^\prime$ refers to the new temporal resolution after aggregation. An example of a temporal processing done in the computation of the Standardized Precipitation Index (SPI), consists of summing the monthly precipitation series over a rolling time window of size $k$. That is also known as the time scale. For SPI, the choice of the time scale $k$ is used to control the accumulation period for the water deficit, enabling the assessment of drought severity across various types (meteorological, agricultural, and hydrological).

@@

The spatial processing module takes a single variable with a fixed temporal dimension, $x_{tj}(s)$, as input. This step transforms the variable from the original spatial dimension $s$ into the new dimension $s^\prime \in D_{s^\prime}$ through $x^{\text{Spat}}_{tj}(s^\prime) = g[x_{tj}(s)]$. The change of spatial dimension allows for the alignment of variables collected from different measurements, such as in-situ stations and satellite imagery, or originating from different resolutions. This also includes the aggregation of variables into different levels, such as city, state, and countries scales.

@@

Variable transformation takes input of a single variable $x_j(s;t)$ and reshapes its distribution using the function $T[.]$ to produce $x^{\text{Trans}}_{j}(s;t)$. When a variable has a skewed distribution, transformations such as log, square root, or cubic root can adjust the distribution towards normality. For example, in Human Development Index (HDI), a logarithmic transformation is applied to the variable Gross National Income per capita (GNI), to reduce its impact on HDI, particularly for countries with high GNI values.

@@

Unlike variable transformation, scaling maintains the distributional shape of the variable. It includes techniques such as centering, z-score standardisation, and min-max standardisation and can be expressed as $[x_{j}(s;t) - \alpha]/\gamma$. In Human Development Index (HDI), the three dimensions (health, education, and economy) are converted into the same scale (0-1) using min-max standardisation.

@@

Distribution fit applies the Cumulative Distribution Function (CDF) $F$ of a distribution on the variable $x_j(s; t)$ to obtain the probability values $p_j(s;t) \in [0, 1]$. In SPEI, many distributions, including log-logistic, pearson III, lognormal, and general extreme distribution, are candidates for the aggregated series. Different fitting methods, and different goodness of fit tests may be used to compare the distribution choice on the index value.

@@

Normalising applies the inverse normal CDF $\Phi^{-1}$ on the input data to obtain the normal density $z_{j}(s;t)$. Normalising can sometimes be confused with the scaling or variable transformation module, which do not involve using a normal distribution to transform the variable. It is arguably whether normalising and distribution fit should be combined or separated into two modules. A decision has been made to separate them into two modules given the different types of output each module present (probability values for distribution fit and normal density values for normalising).

@@

Benchmark sets a value $b_j(s,t)$ for comparing against the original variable $x_j(s;t)$. This benchmark can be a fixed value consistently across space and time or determined by the data through the function $u[x_j(s;t)]$. Once a benchmark being set, observations can be highlighted for adjustments in other modules or can serve targets for monitoring and planning.

@@

Simplification takes a continuous variable $x_j(s;t)$ and categorises it into a discrete set $A_j(s;t) \in \{a_1, a_2, \cdots, a_j\}$ through a piecewise function,

```{=tex}
\begin{equation}
v[x_i(s;t)] = 
\begin{cases}
a_0 & C_1 \leq x^i(s; t) < C_0 \\
a_1 & C_2 \leq x^i(s; t) < C_1 \\
a_2 & C_3 \leq x^i(s; t) < C_2 \\
\cdots \\
a_z & C_z \leq x^i(s; t)
\end{cases}
\end{equation}
```
This is typically used at the end of the index pipeline to simplify the index to communicate to the public the severity of the concept of interest measured by the index. An example of simplification is to map the calculated SPI to four categories: mild, moderate, severe, and extreme drought.

@@

1)  Top-ranked countries with GGGI \> 0.85,
2)  Countries ranked between 57 and 62, with GGGI values from 0.72 to 0.73, and
3)  Low-ranked countries with GGGI \< 0.6.

### 2023-08-17 tidyindex.tex
use j for variable in the subscript instead in notation

\thanks{CONTACT: H. Sherry
Zhang. Email: \href{mailto:huize.zhang@monash.edu}{\nolinkurl{huize.zhang@monash.edu}}. }
\begin{document}
\captionsetup{labelsep=space}
\maketitle
\textsuperscript{1} Department of Econometrics and Business
Statistics, Monash University, Melbourne,
Victoria, Australia\\ \textsuperscript{2} Institute of
Statistics, University of Natural Resources and Life
Sciences, Vienna, Austria\\ \textsuperscript{3} Department of
Mathematical Sciences, BNU-HKBU United International College, Zhuhai,
Guangdong, China
\begin{abstract}
\begin{itemize}
\tightlist
\item
  indexes, useful, quantify severity, early monitoring,
\item
  A huge number of indexes have been proposed by domain experts,
  however, a large majority of them are not being adopted, reused, and
  compared in research or in practice.
\item
  One of the reasons for this is the plenty of indexes are quite complex
  and there is no obvious easy-to-use implementation to apply them to
  user's data.
\item
  The paper describes a general pipeline framework to construct indexes
  from spatio-temporal data,
\item
  This allows all the indexes to be constructed through a uniform data
  pipeline and different indexes to vary on the details of each step in
  the data pipeline and their orders.
\item
  The pipeline proposed aim to smooth the workflow of index construction
  through breaking down the complicated steps proposed by various
  indexes into small building blocks shared by most of the indexes.
\item
  The framework will be demonstrated with drought indexes as examples,
  but appliable in general to environmental indexes constructed from
  multivariate spatio-temporal data
\end{itemize}
\end{abstract}
\begin{keywords}
\def\sep{;\ }
indexes\sep data pipeline\sep 
software design
\end{keywords}
\ifdefined\Shaded\renewenvironment{Shaded}{\begin{tcolorbox}[interior hidden, frame hidden, enhanced, breakable, boxrule=0pt, sharp corners, borderline west={3pt}{0pt}{shadecolor}]}{\end{tcolorbox}}\fi

@@

\begin{equation}
\mathbf{x}(s;t) = \{x^1(s;t), x^2(s;t), \cdots, x^p(s;t)\} \qquad s \in D_s \subseteq \mathbb{R}^m, t \in D_t \subseteq \mathbb{R}^n 
\end{equation}

@@

\begin{itemize}
\item
  \(x^i(s, t)\) represents a variable of interest for example
  precipitation, \(i = 1, \cdots, p\) and
\item
  \(s\) represents the geographic locations in the space
  \(D_s \subseteq \mathbb{R}^m\). Examples of geographic locations
  include a collection of countries, longitude and latitude coordinates
  or regions of interest and,
\item
  \(t\) denotes the temporal order in \(D_t \subseteq \mathbb{R}^n\).
  For instance, time measurements could be recorded hourly, yearly,
  monthly, quaterly or by season.
\end{itemize}

In what follows when geographic or temporal components of the \(x(s,t)\)
process are fixed we will be using suffix notation. For example,
\(x_{s}(t)\) represents the data for a fixed location \(s\) as a
function of time \(t\). While \(x_{t}(s)\) denotes the spatial varying
proccess for a fixed \(t\). An overview of the notation for pipeline
input, operation, and output is present in Table~\ref{tbl-notation}.

\hypertarget{tbl-notation}{}
\begin{longtable}[]{@{}
  >{\raggedright\arraybackslash}p{(\columnwidth - 8\tabcolsep) * \real{0.0128}}
  >{\raggedright\arraybackslash}p{(\columnwidth - 8\tabcolsep) * \real{0.2436}}
  >{\raggedright\arraybackslash}p{(\columnwidth - 8\tabcolsep) * \real{0.2179}}
  >{\raggedright\arraybackslash}p{(\columnwidth - 8\tabcolsep) * \real{0.2436}}
  >{\raggedright\arraybackslash}p{(\columnwidth - 8\tabcolsep) * \real{0.2821}}@{}}
\caption{\label{tbl-notation}An notation overview of the input,
operation, and output of each pipeline module.}\tabularnewline
\toprule()
\begin{minipage}[b]{\linewidth}\raggedright
No.
\end{minipage} & \begin{minipage}[b]{\linewidth}\raggedright
Module
\end{minipage} & \begin{minipage}[b]{\linewidth}\raggedright
Input
\end{minipage} & \begin{minipage}[b]{\linewidth}\raggedright
Operation
\end{minipage} & \begin{minipage}[b]{\linewidth}\raggedright
Output
\end{minipage} \\
\midrule()
\endfirsthead
\toprule()
\begin{minipage}[b]{\linewidth}\raggedright
No.
\end{minipage} & \begin{minipage}[b]{\linewidth}\raggedright
Module
\end{minipage} & \begin{minipage}[b]{\linewidth}\raggedright
Input
\end{minipage} & \begin{minipage}[b]{\linewidth}\raggedright
Operation
\end{minipage} & \begin{minipage}[b]{\linewidth}\raggedright
Output
\end{minipage} \\
\midrule()
\endhead
1 & Temporal processing & \(x^i_{s}(t)\) & \(f[x^i_s(t)]\) &
\(\tilde{x}^i_s(t^\prime)\), where \(t^\prime \in D_{t^\prime}\) \\
2 & Spatial processing & \(x^i_t(s)\) & \(g[x^i_t(s)]\) &
\(x^i_t(s^\prime)\), where \(s^\prime \in D_{s^\prime}\) ???? \\
3 & Variable transformation & \(x^i(s; t)\) & \(T[x^i(s;t)]\) &
\(Tx^i(s;t)\) ???? \\
4 & Scaling & \(x^i(s; t)\) & \([x^i(s;t) - \alpha]/\gamma\) &
\(Sx^i(s;t)\) ???? \\
5 & Dimension reduction & \(\mathbf{x}(s;t)\), where
\(\mathbf{x} \subseteq \mathbb{R}^q,\; q\leq p\) &
\(h[\mathbf{x}(s;t)]\) & \(\mathbf{y}(s;t)\), where
\(\mathbf{y} \subseteq \mathbb{R}^d\) and \(d < q\) \\
6 & Distribution fit & \(x^i(s; t)\) & \(F[x^i(s;t)]\), where \(F\) is
the CDF & \(p^i(s;t)\), where \(p(.) \in [0, 1]\) is a probability
value \\
7 & Normalising & \(x^i(s; t)\) & \(\Phi^{-1}[x^i(s; t)]\) &
\(z^i(s; t)\), where \(z\) has a normal distribution \\
8 & Benchmarking & \(x^i(s; t)\) & \(u[x^i(s;t)]\) & \(b^i(s;t)\) \\
9 & Simplification & \(x^i(s; t)\) & \(v[x^i(s;t)]\) &
\(A^i(s;t) \in \{a_1, a_2, \cdots, a_j\}\) \\
\bottomrule()
\end{longtable}

@@

The temporal processing module takes as input argument a single variable
\(x_s^i(t)\) at location \(s\) as a function of time. In this step the
original time series can be transformed or summarized into a new one via
time aggregation, the transformation is represented by the function
\(f\), \(\tilde{x}_s^i(t^\prime) = f[x^i_s(t)]\) where \(t^\prime\)
refers to the new temporal resolution after aggregation. An example of a
temporal processing done in the computation of the Standardized
Precipitation Index (SPI), consists of summing the monthly precipitation
series over a rolling time window of size \(k\). That is also known as
the time scale. For SPI, the choice of the time scale \(k\) is used to
control the accumulation period for the water deficit, enabling the
assessment of drought severity across various types (meteorological,
agricultural, and hydrological).

@@

The spatial processing module takes a single variable with a fixed
temporal dimension, \(x_t(s)\), as input. This step transforms the
variable from the original spatial dimension \(s\) into the new
dimension \(s^\prime \in D_{s^\prime}\) through
\(x^\#(s;t) = g[x(s;t)]\). The change of spatial dimension allows for
the alignment of variables collected from different measurements, such
as in-situ stations and satellite imagery, or originating from different
resolutions. This also includes the aggregation of variables into
different levels, such as city, state, and countries scales.

@@

Variable transformation takes input of a single variable \(x^i(s;t)\)
and reshapes its distribution using the function \(T[.]\) to produce
\(x^i(s;t)\). When a variable has a skewed distribution, transformations
such as log, square root, or cubic root can adjust the distribution
towards normality. For example, in Human Development Index (HDI), a
logarithmic transformation is applied to the variable Gross National
Income per capita (GNI), to reduce its impact on HDI, particularly for
countries with high GNI values.

@@

Unlike variable transformation, scaling maintains the distributional
shape of the variable. It includes techniques such as centering, z-score
standardisation, and min-max standardisation and can be expressed as
\([x^i(s;t) - \alpha]/\gamma\). In Human Development Index (HDI), the
three dimensions (health, education, and economy) are converted into the
same scale (0-1) using min-max standardisation.

@@

Distribution fit applies the cumulative distribution function \(F\) of a
distribution on the variable \(x^i(s; t)\) to obtain the probability
values \(p^i(s;t) \in [0, 1]\). In SPEI, many distributions, including
log-logistic, pearson III, lognormal, and general extreme distribution,
are candidates for the aggregated series. Different fitting methods, and
different goodness of fit tests may be used to compare the distribution
choice on the index value.

@@

@@

Benchmark sets a value \(b^i(s,t)\) for comparing against the original
variable \(x^i(s;t)\). This benchmark can be a fixed value consistently
across space and time or determined by the data through the function
\(u[x^i(s;t)]\). Once a benchmark being set, observations can be
highlighted for adjustments in other modules or can serve targets for
monitoring and planning.

@@

Simplification takes a continuous variable \(x^i(s;t)\) and categorises
it into a discrete set \(A^i(s;t) \in \{a_1, a_2, \cdots, a_j\}\)
through a piecewise function,

>>>>

\thanks{CONTACT: H. Sherry
Zhang. Email: \href{mailto:huize.zhang@monash.edu}{\nolinkurl{huize.zhang@monash.edu}}. }
\begin{document}
\captionsetup{labelsep=space}
\maketitle
\textsuperscript{1} Department of Econometrics and Business
Statistics, Monash University, Melbourne,
Victoria, Australia\\ \textsuperscript{2} Institute of
Statistics, University of Natural Resources and Life
Sciences, Vienna, Austria\\ \textsuperscript{3} Department of
Mathematical Sciences, BNU-HKBU United International College, Zhuhai,
Guangdong, China
\begin{abstract}
\begin{itemize}
\tightlist
\item
  indexes, useful, quantify severity, early monitoring,
\item
  A huge number of indexes have been proposed by domain experts,
  however, a large majority of them are not being adopted, reused, and
  compared in research or in practice.
\item
  One of the reasons for this is the plenty of indexes are quite complex
  and there is no obvious easy-to-use implementation to apply them to
  user's data.
\item
  The paper describes a general pipeline framework to construct indexes
  from spatio-temporal data,
\item
  This allows all the indexes to be constructed through a uniform data
  pipeline and different indexes to vary on the details of each step in
  the data pipeline and their orders.
\item
  The pipeline proposed aim to smooth the workflow of index construction
  through breaking down the complicated steps proposed by various
  indexes into small building blocks shared by most of the indexes.
\item
  The framework will be demonstrated with drought indexes as examples,
  but appliable in general to environmental indexes constructed from
  multivariate spatio-temporal data
\end{itemize}
\end{abstract}
\begin{keywords}
\def\sep{;\ }
indexes\sep data pipeline\sep 
software design
\end{keywords}
\ifdefined\Shaded\renewenvironment{Shaded}{\begin{tcolorbox}[enhanced, breakable, interior hidden, borderline west={3pt}{0pt}{shadecolor}, frame hidden, sharp corners, boxrule=0pt]}{\end{tcolorbox}}\fi

@@

\begin{equation}
\mathbf{x}(s;t) = \{x_1(s;t), x_2(s;t), \cdots, x_p(s;t)\} \qquad s \in D_s \subseteq \mathbb{R}^m, t \in D_t \subseteq \mathbb{R}^n 
\end{equation}

@@

\begin{itemize}
\item
  \(x_j(s, t)\) represents a variable of interest for example
  precipitation, \(j = 1, \cdots, p\) and
\item
  \(s\) represents the geographic locations in the space
  \(D_s \subseteq \mathbb{R}^m\). Examples of geographic locations
  include a collection of countries, longitude and latitude coordinates
  or regions of interest and,
\item
  \(t\) denotes the temporal order in \(D_t \subseteq \mathbb{R}^n\).
  For instance, time measurements could be recorded hourly, yearly,
  monthly, quaterly or by season.
\end{itemize}

In what follows when geographic or temporal components of the
\(x_j(s,t)\) process are fixed we will be using suffix notation. For
example, \(x_{sj}(t)\) represents the data for a fixed location \(s\) as
a function of time \(t\). While \(x_{tj}(s)\) denotes the spatial
varying process for a fixed \(t\). An overview of the notation for
pipeline input, operation, and output is present in
Table~\ref{tbl-notation}.

\hypertarget{tbl-notation}{}
\begin{longtable}[]{@{}
  >{\raggedright\arraybackslash}p{(\columnwidth - 8\tabcolsep) * \real{0.0294}}
  >{\raggedright\arraybackslash}p{(\columnwidth - 8\tabcolsep) * \real{0.2206}}
  >{\raggedright\arraybackslash}p{(\columnwidth - 8\tabcolsep) * \real{0.1471}}
  >{\raggedright\arraybackslash}p{(\columnwidth - 8\tabcolsep) * \real{0.2206}}
  >{\raggedright\arraybackslash}p{(\columnwidth - 8\tabcolsep) * \real{0.3824}}@{}}
\caption{\label{tbl-notation}An notation overview of the input,
operation, and output of each pipeline module.}\tabularnewline
\toprule()
\begin{minipage}[b]{\linewidth}\raggedright
No.
\end{minipage} & \begin{minipage}[b]{\linewidth}\raggedright
Module
\end{minipage} & \begin{minipage}[b]{\linewidth}\raggedright
Input
\end{minipage} & \begin{minipage}[b]{\linewidth}\raggedright
Operation
\end{minipage} & \begin{minipage}[b]{\linewidth}\raggedright
Output
\end{minipage} \\
\midrule()
\endfirsthead
\toprule()
\begin{minipage}[b]{\linewidth}\raggedright
No.
\end{minipage} & \begin{minipage}[b]{\linewidth}\raggedright
Module
\end{minipage} & \begin{minipage}[b]{\linewidth}\raggedright
Input
\end{minipage} & \begin{minipage}[b]{\linewidth}\raggedright
Operation
\end{minipage} & \begin{minipage}[b]{\linewidth}\raggedright
Output
\end{minipage} \\
\midrule()
\endhead
1 & Temporal processing & \(x_{sj}(t)\) & \(f[x_{sj}(t)]\) &
\(x^{\text{Temp}}_{sj}(t^\prime) \quad t^\prime \in D_{t^\prime}\) \\
2 & Spatial processing & \(x_{tj}(s)\) & \(g[x_{tj}(s)]\) &
\(x^{\text{Spat}}_{tj}(s^\prime) \quad s^\prime \in D_{s^\prime}\) \\
3 & Variable transformation & \(x_{j}(s; t)\) & \(T[x_j(s;t)]\) &
\(x^{\text{Trans}}_j(s;t)\) \\
4 & Scaling & \(x_j(s; t)\) & \([x_j(s;t) - \alpha]/\gamma\) &
\(x^{\text{Scale}}_j(s;t)\) \\
5 & Dimension reduction &
\(\mathbf{x}(s;t) \quad \mathbf{x} \subseteq \mathbb{R}^q,\; q\leq p\) &
\(h[\mathbf{x}(s;t)]\) &
\(\mathbf{y}(s;t) \quad \mathbf{y} \subseteq \mathbb{R}^d, d < q\) \\
6 & Distribution fit & \(x_j(s; t)\) & \(F[x_j(s;t)]\) &
\(p_j(s;t) \quad p(.) \in [0, 1]\) \\
7 & Normalising & \(x_j(s; t)\) & \(\Phi^{-1}[x_j(s; t)]\) &
\(z_j(s; t)\) \\
8 & Benchmarking & \(x_j(s; t)\) & \(u[x_j(s;t)]\) & \(b_j(s;t)\) \\
9 & Simplification & \(x_j(s; t)\) & \(v[x_j(s;t)]\) &
\(A_j(s;t) \in \{a_1, a_2, \cdots, a_j\}\) \\
\bottomrule()
\end{longtable}

@@

The temporal processing module takes as input argument a single variable
\(x_{sj}(t)\) at location \(s\) as a function of time. In this step the
original time series can be transformed or summarized into a new one via
time aggregation, the transformation is represented by the function
\(f\), \(x^{\text{Temp}}_{sj}(t^\prime) = f[x_{sj}(t)]\) where
\(t^\prime\) refers to the new temporal resolution after aggregation. An
example of a temporal processing done in the computation of the
Standardized Precipitation Index (SPI), consists of summing the monthly
precipitation series over a rolling time window of size \(k\). That is
also known as the time scale. For SPI, the choice of the time scale
\(k\) is used to control the accumulation period for the water deficit,
enabling the assessment of drought severity across various types
(meteorological, agricultural, and hydrological).

@@

The spatial processing module takes a single variable with a fixed
temporal dimension, \(x_{tj}(s)\), as input. This step transforms the
variable from the original spatial dimension \(s\) into the new
dimension \(s^\prime \in D_{s^\prime}\) through
\(x^{\text{Spat}}_{tj}(s^\prime) = g[x_{tj}(s)]\). The change of spatial
dimension allows for the alignment of variables collected from different
measurements, such as in-situ stations and satellite imagery, or
originating from different resolutions. This also includes the
aggregation of variables into different levels, such as city, state, and
countries scales.

@@

Variable transformation takes input of a single variable \(x_j(s;t)\)
and reshapes its distribution using the function \(T[.]\) to produce
\(x^{\text{Trans}}_{j}(s;t)\). When a variable has a skewed
distribution, transformations such as log, square root, or cubic root
can adjust the distribution towards normality. For example, in Human
Development Index (HDI), a logarithmic transformation is applied to the
variable Gross National Income per capita (GNI), to reduce its impact on
HDI, particularly for countries with high GNI values.

@@

Unlike variable transformation, scaling maintains the distributional
shape of the variable. It includes techniques such as centering, z-score
standardisation, and min-max standardisation and can be expressed as
\([x_{j}(s;t) - \alpha]/\gamma\). In Human Development Index (HDI), the
three dimensions (health, education, and economy) are converted into the
same scale (0-1) using min-max standardisation.

@@

Distribution fit applies the Cumulative Distribution Function (CDF)
\(F\) of a distribution on the variable \(x_j(s; t)\) to obtain the
probability values \(p_j(s;t) \in [0, 1]\). In SPEI, many distributions,
including log-logistic, pearson III, lognormal, and general extreme
distribution, are candidates for the aggregated series. Different
fitting methods, and different goodness of fit tests may be used to
compare the distribution choice on the index value.

@@

Normalising applies the inverse normal CDF \(\Phi^{-1}\) on the input
data to obtain the normal density \(z_{j}(s;t)\). Normalising can
sometimes be confused with the scaling or variable transformation
module, which do not involve using a normal distribution to transform
the variable. It is arguably whether normalising and distribution fit
should be combined or separated into two modules. A decision has been
made to separate them into two modules given the different types of
output each module present (probability values for distribution fit and
normal density values for normalising).

@@

Benchmark sets a value \(b_j(s,t)\) for comparing against the original
variable \(x_j(s;t)\). This benchmark can be a fixed value consistently
across space and time or determined by the data through the function
\(u[x_j(s;t)]\). Once a benchmark being set, observations can be
highlighted for adjustments in other modules or can serve targets for
monitoring and planning.

@@

Simplification takes a continuous variable \(x_j(s;t)\) and categorises
it into a discrete set \(A_j(s;t) \in \{a_1, a_2, \cdots, a_j\}\)
through a piecewise function,
